Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-01-24 analysis/paper/paper.Rmd
edit the background and working on the discussion section

<!--the major finding at GJB-->
Among the various ceramic styles, ceramics with surface decorations such as incised geometric patterns and impressed dot patterns are the two common styles that have also been found in western Yunnan, indicative of frequent interactions between Guijiabao and sites in Yunnan [@Guijiabao2017a]. The archaeobotanical remains reveal evidence of crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on the analysis of flotation samples from ash pits and stratigraphic layers in 2015 [@Guijiabao2016]. The adoption of millet cultivation can be traced back to ca. 5000 BP [@Hao2022]. In addition, rice grains have also been found in ash pits throughout different phases where the earliest rice dates to 2881-2633 ca. BP [@Hao2022].

<!--info about Gaoshan-->
Gaoshan site is located in the southwestern margin of the Chengdu Plain, an alluvial plain between the Sichuan Basin and the Tibetan Plateau (Figure \@ref(fig:SW-china-map)). Excavations at Gaoshan revealed abundant pottery, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015]. According to the pottery forms and decorations, Gaoshan can be categorized into the earliest phase of the Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain [@Guedes2013]. It is proposed that Gaoshan could be even older than the Baodun culture according to a burial context, which was found in the stratigraphic layer earlier than the Baodun period [@Gaoshan2017].

A large quantity of carbonized rice grains was found at Gaoshan with a relatively small number of foxtail/broomcorn millets, suggesting rice is the main crop supplemented by millets [@JiangGaoshan2017]. This also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China occurred at least 4500 years ago [@Chi2010]. Moreover, recent research on animal bones in 2015 and 2016 shows that pigs and dogs were already domesticated, of which pigs are the main domestic animal [@He2020]. The primary subsistence strategy at Gaoshan could be rice/millet cultivation with domesticated pigs.

It is generally accepted that mixed farming of millet and rice spread to southwest Sichuan around 5,000 years ago and to western Yunnan before 4600 years ago [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010]. The current finding at Guijiabao and Gaoshan support this argument. However, the ratio of rice to millet grains found at Guijiabao suggests that millet is the major crop in an earlier period, which is different from a clear predominance of rice (*Oryza sativa*) at Gaoshan or other sites in Sichuan and Yunnan, such as Baodun, Henglanshan and Baiyangcun [@Hao2022; @Guedes2013]. Although the analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, evidence directly reflects food preparation or processing using ceramics still lacks. This study provides direct evidence from food residue in ceramics to investigate food choices associated with new practices of crop cultivation and local subsistence.

@@

# Discussion
Our preliminary results show that the food categories indicated by potsherds at Guijiabao and Gaushan could be mainly dominated by terrestrial animals, such as ruminant animals. Although the yield of lipids in most of our potsherd samples is generally scarce, one potsherd demonstrates a relatively rich composition of long-chain fatty acids which confirms their usage as a cooking or storage pot. A larger amount of C16 and C18 fatty acids compared to other types of fatty acids indicate that the organic residues in our potsherds were highly degraded over the past 4500 years. It is also striking to find that one potsherd of Guijiabao contains residues that might originate from C3 plants.

show a mixture of C3 plants, such as rice, and terrestrial animals for most pots with one pot showing more C4 plant source or marine source. Our striking finding might suggest that consumption of rice could be part of the daily life at Guijiabao and the adoption of rice agriculture could be as important as millet cultivation which was thought the main crop previously. Our next steps is to increase sample size to include pots from different time periods and different types of pottery to identify any variation through time and also try different methods, such as phytolith analysis to further support our statement about rice cultivation.

@@

>>>>

<!--the major finding at GJB-->
Among the various ceramic styles, ceramics with surface decorations such as incised geometric patterns and impressed dot patterns are the two common styles that have also been found in western Yunnan, indicative of frequent interactions between Guijiabao and sites in Yunnan [@Guijiabao2017a]. The archaeobotanical remains reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on the analysis of flotation samples from ash pits and stratigraphic layers in 2015 [@Guijiabao2016]. The adoption of millet cultivation can be traced back to ca. 5000 BP [@Hao2022]. In addition, rice grains (*Oryza sativa*) have also been found in ash pits throughout different phases where the earliest rice dates to 4700 ca. BP [@Hao2022]. A detailed phytolith analysis for soil samples dating to Neolithic period (5000 ca. BP to 3700 ca. BP) further suggests the presence of rice phytoliths throughout the whole period in addition to millet phytoliths, indicative of a mixed millet/rice cultivation prior to 5000 ca. BP than previously thought [@Huan2022].

<!--info about Gaoshan-->
Gaoshan site is located in the southwestern margin of the Chengdu Plain, an alluvial plain between the Sichuan Basin and the Tibetan Plateau (Figure \@ref(fig:SW-china-map)). Excavations at Gaoshan revealed abundant pottery, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015]. According to the pottery forms and decorations, Gaoshan can be categorized into the earliest phase of the Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain [@Guedes2013]. It is proposed that Gaoshan could be even older than the Baodun culture according to a burial context, which was found in the stratigraphic layer earlier than the Baodun period [@Gaoshan2017]. A large quantity of carbonized rice grains was found at Gaoshan with a relatively small number of foxtail/broomcorn millets, suggesting rice is the main crop supplemented by millets [@JiangGaoshan2017]. This also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China occurred at least 4500 years ago [@Chi2010]. Moreover, recent research on animal bones in 2015 and 2016 shows that pigs and dogs were already domesticated, of which pigs are the main domestic animal [@He2020]. The primary subsistence strategy at Gaoshan could be rice/millet cultivation with domesticated pigs.

Despite of a long-standing debate between waves of rice/millet dispersal from northwest China and from the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP. and to western Yunnan before 4600 BP. [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010]. The current finding at Guijiabao and Gaoshan support this argument. However, plant remains from Guijiabao shows that millet is the major crop, which is at odds with the predominance of rice based on phytolich analysis and other macro-botanical analyses from Gaoshan or sites in Sichuan and Yunnan, such as Baodun, Henglanshan and Baiyangcun [@Hao2022; @Guedes2013]. Although the analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, evidence directly reflects food preparation or processing using ceramics still lacks. This study aims to provide direct evidence from food residues in ceramics to investigate food choices associated with new practices of crop cultivation and local subsistence.

@@

# Discussion
<!--A brief summary of interesting results-->
Our results show that food categories indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by terrestrial animals, particularly ruminant fats or mixtures of terrestrial animals and plants. Moreover, it is striking to identify a non-ruminant source from a Guijiabao potshard that might be C3 plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields for all potsherds support that those ceramics were used as cooking or storage pots. One potshard dating to around 4500 years ago demonstrates a rich composition of long-chain fatty acids, suggesting a better condition for preserving lipid molecules. We also recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes from both Guijiabao and Gaoshan. However, the general pattern of the isotopic values of lipid molecules could provide informative clues for a better understanding of the local subsistence of Guijiabao.

Ruminant fats in the Neolithic potsherds around ca. 4500 BP. from Guijiabao and Gaoshan could be derived from deer or horse according to the finding of faunal remains, accompanied with stone artifacts for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a]. However, whether horses were part of local food choices in an earlier period is questionable. First, the Pleistocene and Neolithic sites with wild horse bones mostly concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004]. Although wild Asian water buffalo could be another food source due to their presence since the Pleistocene [@Zhang2016], no cattle bones were found at Guijiabao or Gaoshan. Therefore, it is more likely that deer could be the main resource of ruminant fats. The analysis of faunal remains from Guijiabao is still in progress which could allow us to evaluate our findings in the future.

check isotopic results for horse, deer and cattle

Medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded over the past 4500 years.

The results show a mixture of C3 plants, such as rice, and terrestrial animals for most pots with one pot showing more C4 plant source or marine source. Our striking finding might suggest that consumption of rice could be part of the daily life at Guijiabao and the adoption of rice agriculture could be as important as millet cultivation which was thought the main crop previously. Our next steps is to increase sample size to include pots from different time periods and different types of pottery to identify any variation through time and also try different methods, such as phytolith analysis to further support our statement about rice cultivation.

@@

A large-scale analysis of ceramic across the region may offer insight into regional variations in food choices and the spread of agriculture.
